Tools and Ideas for the Procedural Generation of Shadow Sculptures

### 2015-12-16 paper/paper.tex
Some corrections made to the paper

%opening
\title{Tools and Ideas for the Proceedural Generation of Shadow Sculptures.}
\author{James Hoctor, Chris Chappell}

@@

\section*{Introduction}
This paper addresses the proceedural generation of artwork, if such
a thing is even possible. We wish to create art in the style of
Larry Kagan, whose medium is metal and the shadows it casts. In a
typical Kagan piece there is a metal object made from steel bars
that have been bent to shape and welded together to form a whole.
This assembly is mounted on a wall (painted white) and a spot light
shines on it to cast a shadow on the wall that forms an image. In a
naive sense, the problem  of creating something like Kagan's work is
a very simple one: one solution is to create a metal object in the shape of the desired
shadow but scaled down by half, then mount this half way between the
light source and the wall. The mounting can be done by means of
straight bars that travel in the shadow from the rest of the object
to the wall. But this is not the kind of result we desire. A major
challenge of this project is determining the other factors that make
a Kagan piece interesting beyond the mechanical fact that an image
is present in the form of a shadow. In particular, the metal portion
of a genuine Kagan is often visually chaotic and does not contain
recognizable forms. (Exceptions to this are discussed in the section
\ref{kagan}.) Our approach to generating artwork in this style
is therefore to use a genetic algorithm to search for a configuration
of metal that meets a variety of conditions among which is the
obvious requirement that the shadow appear as desired.

Other components of fitness of the object in the genetic algorithm
are its structural stability, its visual chaoticness, and its ability
to be fabricated using techniques similar to what Kagan uses. Structural stability
is addressed in section \ref{GA}. Visual
chaos is achieved by the random nature of the selection process.
Fabricability is included because it adds another constraint in a
problem that has few. It was key consideration when we decided
how to represent these sculptures.

\section{Larry Kagan's Work} \label{kagan}
In researching this problem we made a brief study of Kagan's work for
several reasons. Firstly, we wanted to identify a subset of his work
that might be most easily imitated. Secondly, our analysis shows how
necessary this restriction is. Kagan's work is too varied for the
whole body of work to be imitated. There is also the question of how
Kagan selects the image he wishes to create from shadow. In all of our
work we leave this crucial piece of creative heavy lifting to the
experts: humans.

Each of Larry Kagan's sculptures consists of a metal structure
composed of bent rods of metal mounted to a wall and illuminated by
a bright light from a carefully chosen location. The eye is drawn to
the shadow produced this way, which forms a familiar image, while the
rods themselves often have no obvious pattern. In the simplest case,
the shadow image is a line drawing and the rods are attached to the
wall at several points. For example, the piece ``Frankfurt Chair'' is
like this. Kagan has created many variations on this idea including
incorporating some of the bent rods in the recognizable image while
most of the image is formed by the shadow, as seen in ``Beach Chair''.
``Intersection of Two Circles'' shows that a sculpture composed of rods
can have a solid shadow with a geometric shape. The light source can
be on the ceiling cented over the piece or be off to one side as it
is in ``At''. With the light source on one side the shadow falls
predominantly on the side of the sculpture. Despite the fact that a
single small bright light source is installed to cast the shadow for
each sculpture, soft egdes of the shadow are noticeable in many of
these works. In ``Mosquito I'', Kagan uses this effect in a dramatic
way; the image present in the shadow of this sculpture is a mosquito
with its own shadow cast by an imaginary light source. In order to
make the distinction obvious, Kagan allowed soft shadow edges to show
in the portion of the image corresponding to the imaginary shadow of
the mosquito, but created the image of the mosquito with dark
hard-edged shadows. All of these sculptures (with the exception of ``At'') can be viewed on Larry Kagan's website \cite{Kagan}.

@@

\section{How We Render Sculptures} \label{sculpture_render}
Using the method of spheres and cylinders, objects can be graphically modeled and displayed in 3D rendering systems fairly easily.  One method of doing this that we tested was using OpenGL to build a little engine that could render the object and a small surface such that the object and the shadow could be viewed from multiple angles, as it would look if it were printed.  The shadow could be imported as a texture, using OpenGL's \texttt{GL\_TEXTURE2D} object, and then fitted to the surface that would compose the wall.  This would allow modularity and would separate the methods that composed the creation of the shadow and its rendering.  While this may not be optimal, it is the most easily supported method using OpenGL legacy code.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{object-render3.png}
    \caption{A sample render of ten spheres and their shadows.  The light source used was made up of nine point light sources, grouped together very closely.}
    \label{fig:render_3}
\end{figure}

To further explain the algorithm behind this method, first let us describe how the object would be represented.  The method used is most similar to the second representation of objects discussed previously, as spheres and cylinders in a graph layout.  Each object would consist of a list of nodes or spheres $W$, a list of edges or cylinders $E$, and a radius $r$.  The nodes in $W$ are described using 3-tuples of floating point numbers, representing their location in Cartesian 3-space, and the edges in $E$ are represented by 2-tuples of indexes, referencing the two nodes in $W$.  The algorithm iterates over the nodes in $W$, creating spheres of radius $r$.  Then, it iterates over the edges in $E$, and for each edge $e$ creates cylinders with radius $r$ and endpoints centered on the two nodes in $W$ indexed by $e$.  This is done using pyOpenGL's \texttt{gluSphere()} to draw spheres and \texttt{gluCylinder()} to draw cylinders.  For more information on the programming in this implementation, see \texttt{renderObject.py}.\cite{gitrepo}

For a more updated model that could possibly fix both the issue of slow shadow rendering time and separation of the shadow rendering from the object's rendering, one may instead of legacy OpenGL use OpenGL 3.3 or above, which adds the use of fragment shaders, and can render shadows from point light sources fairly easily.  The question that arises in this method, however, is how one would compare two shadows to determine accuracy and fitness of an object.  This would require further research and testing.

There are many video game engines that have builtin methods for shadow generation, with many options and customizable settings for doing so.  This may be the simplest option for rendering objects, and possibly even would also allow for comparison and heuristic generation.  Such engines would include Unity, Unreal Engine, and Source 2, just to name a few.  As our knowledge did not extend to the realm of writing video games, we felt more comfortable working with OpenGL.

@@

for each \(sphere\) in \(spheres\):
  if the sphere is entirely above the light or below the shadow plane:
    it casts no shadow; continue to the next sphere
  if the top of the sphere is beneath the light and the bottom is above the plane:
    The shadow cast on the plane will be elliptical.
    Calculate the equation of the cone formed by all lines passing through the
        light source and tangent to the sphere.
    Find the equation of the ellipse by intersecting this cone with the shadow
        plane.
    The equation has the form \(A*x^2 + B*y^2 + C*x*y + D*x + E*y + F = 0\).
    Rearrange the equation to have the form \(G(x)*y^2 + H(x)*y + I(x) = 0\), where
        \(G\), \(H\), and \(I\) are 2nd order or lower polynomials in x.
    For each column \(c\) in pixels:
      Find the value of \(x\) that corresponds to \(c\).
      Calculate \(G(x)\), \(H(x)\), and \(I(x)\).
      Use the quadratic formula to find two values of \(y\) satisfying the
          equation for the boundary of the ellipse.
      Convert the values of \(y\) to row coordinates in the image.
      In column \(c\) and all rows \(r\) between the two calculated:
        color \(pixels[r][c]\) black
  if the sphere intersects the shadow plane:
    Handling this case is at least as complicated as handling a sphere strictly
        above the plane.
    The circular intersection of the circle and the plane is shadowed.
    Any additional shadowing is a subset of the elliptical shadow described in
        the previous case.
    This case is not handled by our code.
  if the sphere intersects the light:
    Color all of pixels black.
    Return from the proceedure early.
    This case is not handled by our code.
  if the top of the sphere is above the light, its bottom below the light, without
      intersecting it:
    The shadow cast is hyperbolic.
    This case is not handled by our code, but it would likely be similar to
        the case of elliptical shadows.

@@

We did not implement casting the shadows of cylinders by this method, but there is an optimization that deserves mention. It results from the fact that we can assume there is a sphere present at each end of any cylinder. Consequentially, any cylinder that is aligned with one face toward the light source casts a shadow only where a sphere has already cast a shadow. The optimization that arises from this is to ignore cylinders with such an orientation in the loop over all cylinders.

@@

For each light source we imagine a rectangular pyramid with the region-of-interest of the shadow plane as its base and the light source as its vertex. (See figure \ref{fig:space_partitioning}). We divide the pyramid into chunks and to each chunk associate all the spheres with a center in the chunk. When checking if a line segment intersects any sphere, we can ignore any sphere not in a chunk near the line segment. In particular, we started by dividing the pyramid into four levels. At the top level there are no subdivisions as we would have to check the spheres in each division anyway. This is because any partition that gets arbitrarity close to the light source can include spheres that shadow the entire shadow plane. The next level down is divided into truncated pyramids. The top plane of this layer can be thought of as divided into squares with a side length equal to $r$ the radius of the spheres. A ray of light from the source will pass through one of these squares on its way to the shadow plane unless it is blocked in the previous layer. In this second highest layer it can only be blocked by spheres whose centers are beneath that square or any adjacent square. Therefore, we check for intersection only with spheres that fall in these partitions. In python we implement this by finding the intersection of the light path with the top plane and calculating integer coordinates of the square in which the intersection falls. Then from this coordinate pair is calculated a list of adjacent coordinates. All of these pairs are used in turn to perform a look-up operation in a python dictionary object each of which results in a list of spheres. Only these spheres are checked for intersection. The levels beneath that one work the same way. The benefit to dividing the lower part of the pyramid further is that lower levels have a greater capacity to divide the spheres. As noted in the annotations on figure \ref{fig:space_partitioning}, a sphere will be sorted into a partition in a level higher than its center if it intersects the boundary between the layers. This is necessary because the part of the sphere above the plane dividing the levels might block a light path that does not go through an adjacent square. The problem is avoided by putting the sphere in the higher level. It might seem at first that this trick makes the method no longer a space-partioning method, but it fact it merely means that the partitions are not as drawn in figure \ref{fig:space_partitioning}. Instead the top and bottom of a partition is at a distance $r$ below the planes drawn in figure \ref{fig:space_partitioning} (With the exception of the top of the top partition and the bottoms of the bottom partitions).

@@

\begin{appendices}
\section{A} \label{app:ellipse_derivation}

@@

\end{appendices}
% ftp://ftp.ams.org/ams/amsrefs/amsrdoc.pdf for amsrefs documentation
\begin{bibdiv}
 \begin{biblist}
  \bib{Dasgupta}{book}{
   title={Algorithms},
   author={Dasgupta, Sanjoy},
   author={Papadimitriou, Christos},
   author={Vazirani, Umesh},
   publisher={McGraw-Hill Education},
   address={New York},
   year={2006}  
  }
  \bib{pyOpenGL}{webpage}{
   title={PyOpenGL Documentation},
   url={http://pyopengl.sourceforge.net/documentation/},
   accessdate={2015-10}
  }
  \bib{gitrepo}{webpage}{
   title={Genetic Algorithms for Shadow Sculptures},
   author={Chris Chappell},
   author={James Hoctor},
   url={https://github.com/chappc/Math-Senior-Research},
   accessdate={2015-12}
  }
  \bib{cylindermagic}{webpage}{
   title={Rendering a Cylinder Between Two Points in OpenGL},
   author={Parris, Joel},
   url={http://lifeofaprogrammergeek.blogspot.com/2008/07/rendering-cylinder-between-two-points.html},
   accessdate={2015-10}
  }
  \bib{opengl-tutorial}{webpage}{
   title={opengl-tutorial},
   url={http://www.opengl-tutorial.org/},
   accessdate={November 2015}
  }
  \bib{Kagan}{webpage}{
   title={Larry Kagan Sculpture},
   url={http://larrykagansculpture.com/},
   accessdate={October 2015}
  }
  \bib{GLM}{webpage}{
   title={OpenGL Mathematics Library},
   url={http://glm.g-truc.net/0.9.4/index.html},
   accessdate={December 2015}
  }
 \end{biblist}

>>>>

%opening
\title{Tools and Ideas for the Procedural Generation of Shadow Sculptures}
\author{James Hoctor, Chris Chappell}

@@

\section*{Introduction}
This paper addresses the procedural generation of artwork, if such
a thing is even possible. We wish to create art in the style of
Larry Kagan, whose medium is metal and the shadows it casts. In a
typical Kagan piece there is a metal object made from steel bars
that have been bent to shape and welded together to form a whole.
This assembly is mounted on a white wall and a spot light
shines on it to cast a shadow that forms an image on the wall. In a
naive sense, the problem  of creating something like Kagan's work is
a very simple one: one solution is to create a metal object in the shape of the desired
shadow but scaled down by half, then mount this half way between the
light source and the wall. The mounting can be done by means of
straight bars that travel in the shadow from the rest of the object
to the wall. But this is not the kind of result we desire. A major
challenge of this project is determining the other factors that make
a Kagan piece interesting beyond the mechanical fact that an image
is present in the form of a shadow. In particular, the metal portion
of a genuine Kagan is often visually chaotic and does not contain
recognizable forms. (Exceptions to this are discussed in the section
\ref{kagan}.) Our approach to generating artwork in this style
is therefore to use a genetic algorithm to search for a configuration
of metal that meets a variety of conditions among which is the
obvious requirement that the shadow appear as desired.

Other components of fitness of the object in the genetic algorithm
are its structural stability, its visual chaoticness, and its ability
to be fabricated using techniques similar to what Kagan uses. Structural stability
is addressed in section \ref{GA}. Visual
chaos is achieved by the random nature of the selection process.
Fabricability is included because it adds another constraint in a
problem that has few. It was a key consideration when we decided
how to represent these sculptures.

\section{Larry Kagan's Work} \label{kagan}
In researching this problem we studied of Kagan's work for
several reasons. Firstly, we wanted to identify a subset of his work
that might be most easily imitated. Secondly, our analysis shows how
necessary this restriction is. Kagan's work is too varied for the
whole body of work to be imitated. There is also the question of how
Kagan selects the image he wishes to create from shadow. In all of our
work we leave this crucial piece of creative heavy lifting to the
experts: humans.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.8\textwidth]{mosquito_full.jpg}
    \caption{Larry Kagan's ``Mosquito I''}
    \label{fig:mosquito}
\end{figure}

Each of Kagan's sculptures consists of a metal structure
composed of bent rods of metal mounted to a wall and illuminated by
a bright light from a carefully chosen location. The eye is drawn to
the shadow produced this way, which forms a familiar image, while the
rods themselves often have no obvious pattern. In the simplest case,
the shadow image is a line drawing and the rods are attached to the
wall at several points. For example, the piece ``Frankfurt Chair'' is
like this. Kagan has created many variations on this idea including
incorporating some of the bent rods in the recognizable image while
most of the image is formed by the shadow, as seen in ``Beach Chair''.
``Intersection of Two Circles'' shows that a sculpture composed of rods
can have a solid shadow with a geometric shape. The light source can
be on the ceiling cented over the piece or be off to one side as it
is in ``At''. With the light source on one side the shadow falls
predominantly on the side of the sculpture. Despite the fact that a
single small bright light source is installed to cast the shadow for
each sculpture, soft egdes of the shadow are noticeable in many of
these works. In ``Mosquito I'' \ref{fig:mosquito}, Kagan uses this effect in a dramatic
way; the image present in the shadow of this sculpture is a mosquito
with its own shadow cast by an imaginary light source. In order to
make the distinction obvious, Kagan allowed soft shadow edges to show
in the portion of the image corresponding to the imaginary shadow of
the mosquito, but created the image of the mosquito with dark
hard-edged shadows. All of these sculptures can be viewed on Larry Kagan's website \cite{Kagan}.

@@

\section{How We Render Sculptures} \label{sculpture_render}
Using the method of spheres and cylinders, objects can be graphically modeled and displayed in 3D rendering systems fairly easily.  One method that we tested was using OpenGL to build a little engine that could render the object and a small surface, such that the object and the shadow could be viewed from multiple angles.  The shadow could be imported as a texture, using OpenGL's \texttt{GL\_TEXTURE2D} object, and then fitted to the surface that would compose the wall.  This would allow modularity and would separate the methods that composed the creation of the shadow and its rendering.  While this may not be optimal, it is the most easily supported method using OpenGL legacy code.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.8\textwidth]{object-render3.png}
    \caption{A sample render of ten spheres and their shadows.  The light source used was made up of nine point light sources, grouped together very closely.}
    \label{fig:render_3}
\end{figure}

To further explain the algorithm behind this method, first let us describe how the object would be represented.  The method used is most similar to the second representation of objects discussed previously, as spheres and cylinders in a graph layout.  Each object would consist of a list of nodes or spheres $W$, a list of edges or cylinders $E$, and a radius $r$.  The nodes in $W$ are described using 3-tuples of floating point numbers, representing their location in Cartesian 3-space, and the edges in $E$ are represented by 2-tuples of indices, referencing the two nodes in $W$.  The algorithm iterates over the nodes in $W$, creating spheres of radius $r$.  Then, it iterates over the edges in $E$, and for each edge $e$ creates cylinders with radius $r$ and endpoints centered on the two nodes in $W$ indexed by $e$.  This is done using pyOpenGL's \texttt{gluSphere()} to draw spheres and \texttt{gluCylinder()} to draw cylinders.  For more information on the programming in this implementation, see \texttt{renderObject.py}.\cite{gitrepo}

For a more updated model that could potentially fix both the issue of slow shadow rendering time and separation of the shadow rendering from the object's rendering, one may instead of legacy OpenGL use OpenGL 3.3 or above.  OpenGL 3.3 adds the use of fragment shaders, and can render shadows from point light sources fairly easily.  The question that arises in this method, however, is how one would compare two shadows to determine accuracy and fitness of an object.  This would require further research and testing.

There are many video game engines that have builtin methods for shadow generation, with many options and customizable settings for doing so.  This may be the simplest option for rendering objects, and potentially would allow for comparison and heuristic generation.  Such engines would include Unity, Unreal Engine, and Source 2, just to name a few.  As our knowledge did not extend to the realm of writing video games, we felt more comfortable working with OpenGL.

@@

for each \(sphere\) in \(spheres\):
  if the sphere is entirely above the light or below the shadow plane:
    it casts no shadow; continue to the next sphere.
  if the top of the sphere is beneath the light and the bottom is above the plane:
    The shadow cast on the plane will be elliptical.
    Calculate the equation of the cone formed by all lines passing through the
        light source and tangent to the sphere.
    Find the equation of the ellipse by intersecting this cone with the shadow
        plane.
    The equation has the form \(A*x^2 + B*y^2 + C*x*y + D*x + E*y + F = 0\).
    Rearrange the equation to have the form \(G(x)*y^2 + H(x)*y + I(x) = 0\), where
        \(G\), \(H\), and \(I\) are 2nd order or lower polynomials in x.
    For each column \(c\) in pixels:
      Find the value of \(x\) that corresponds to \(c\).
      Calculate \(G(x)\), \(H(x)\), and \(I(x)\).
      Use the quadratic formula to find two values of \(y\) satisfying the
          equation for the boundary of the ellipse.
      Convert the values of \(y\) to row coordinates in the image.
      In column \(c\) and all rows \(r\) between the two calculated:
        color \(pixels[r][c]\) black.
  if the sphere intersects the shadow plane:
    Handling this case is at least as complicated as handling a sphere strictly
        above the plane.
    The circular intersection of the circle and the plane is shadowed.
    Any additional shadowing is a subset of the elliptical shadow described in
        the previous case.
    This case is not handled by our code.
  if the sphere intersects the light:
    Color all of pixels black.
    Return from the proceedure early.
    This case is not handled by our code.
  if the top of the sphere is above the light, its bottom below the light, without
      intersecting it:
    The shadow cast is hyperbolic.
    This case is not handled by our code, but it would likely be similar to
        the case of elliptical shadows.

@@

We did not implement casting the shadows of cylinders by this method, but nonetheless there is an optimization that deserves mention in regards to casting shadows of cylinders. It results from the fact that we can assume there is a sphere present at each end of any cylinder. Consequentially, any cylinder that is aligned with one face toward the light source casts a shadow only where a sphere has already cast a shadow. The optimization that arises from this is to ignore cylinders with such an orientation in the loop over all cylinders.

@@

For each light source we imagine a rectangular pyramid with the region-of-interest of the shadow plane as its base and the light source as its vertex. (See figure \ref{fig:space_partitioning}). We divide the pyramid into chunks and to each chunk associate all the spheres with a center in the chunk. When checking if a line segment intersects any sphere, we can ignore any sphere not in a chunk near the line segment. In particular, we started by dividing the pyramid into four levels. At the top level there are no subdivisions as we would have to check the spheres in each division anyway. This is because any partition that gets arbitrarity close to the light source can include spheres that shadow the entire shadow plane. The next level down is divided into truncated pyramids. The top plane of this layer can be though of as divided into squares with a side length equal to $r$ the radius of the spheres. A ray of light from the source will pass through one of these squares on its way to the shadow plane unless it is blocked in the previous layer. In this second highest layer it can only be blocked by spheres whose centers are beneath that square or any adjacent square. Therefore, we check for intersection only with spheres that fall in these partitions. In python we implement this by finding the intersection of the light path with the top plane and calculating integer coordinates of the square in which the intersection falls. Then from this coordinate pair is calculated a list of adjacent coordinates. All of these pairs are used in turn to perform a look-up operation in a python dictionary object each of which results in a list of spheres. Only these spheres are checked for intersection. The levels beneath that one work the same way. The benefit to dividing the lower part of the pyramid further is that lower levels have a greater capacity to divide th spheres. As noted in the annotations on figure \ref{fig:space_partitioning}, a sphere can be sorted into a partition in a level higher than its center if it intersects the boundary between the layers. This is necessary because the part of the sphere above the plane dividing the levels might block a light path that does not go through an adjacent square. The problem is avoided by putting the sphere in the higher level. It might seem at first that this trick makes the method no longer a space-partioning method, but it fact it merely means that the partitions are not as drawn in figure \ref{fig:space_partitioning}. Instead the top and bottom of a partition is at a distance $r$ below the planes drawn in figure \ref{fig:space_partitioning} (With the exception of the top of the top partition and the bottoms of the bottom partitions).

@@

\begin{appendices}
\section{Shadow Calculations} \label{app:ellipse_derivation}

@@

\end{appendices}
% ftp://ftp.ams.org/ams/amsrefs/amsrdoc.pdf for amsrefs documentation
\begin{bibdiv}
 \begin{biblist}
  \bib{Dasgupta}{book}{
   title={Algorithms},
   author={Dasgupta, Sanjoy},
   author={Papadimitriou, Christos},
   author={Vazirani, Umesh},
   publisher={McGraw-Hill Education},
   address={New York},
   year={2006}  
  }
  \bib{pyOpenGL}{webpage}{
   title={PyOpenGL Documentation},
   url={http://pyopengl.sourceforge.net/documentation/},
   accessdate={2015-10}
  }
  \bib{gitrepo}{webpage}{
   title={Genetic Algorithms for Shadow Sculptures},
   author={Chappell, Chris},
   author={Hoctor, James},
   url={https://github.com/chappc/Math-Senior-Research},
   accessdate={2015-12}
  }
  \bib{cylindermagic}{webpage}{
   title={Rendering a Cylinder Between Two Points in OpenGL},
   author={Parris, Joel},
   url={http://lifeofaprogrammergeek.blogspot.com/2008/07/rendering-cylinder-between-two-points.html},
   accessdate={2015-10}
  }
  \bib{opengl-tutorial}{webpage}{
   title={opengl-tutorial},
   url={http://www.opengl-tutorial.org/},
   accessdate={November 2015}
  }
  \bib{Kagan}{webpage}{
   title={Larry Kagan Sculpture},
   url={http://larrykagansculpture.com/},
   accessdate={October 2015}
  }
  \bib{GLM}{webpage}{
   title={OpenGL Mathematics Library},
   url={http://glm.g-truc.net/0.9.4/index.html},
   accessdate={December 2015}
  }
 \end{biblist}
